---
title: "Parameterize hatchR Models"
author: "Morgan Sparks, Bryan M. Maitland"
bibliography: '`r system.file("references.bib", package="hatchR")`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameterize hatchR Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# rmd style
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE, 
  fig.align = 'center',
  fig.width = 6
)
options(tibble.print_min = 5, tibble.print_max = 5)

# load packages
library(hatchR)
library(tibble)
library(ggplot2)
```

# Overview

This vignette describes the process of parameterizing models for predicting fish phenology using **hatchR**. The package provides two options for choosing phenological models:

1.  use model parameterizations included in the package
2.  create custom parameterizations using your own data

First, we load **hatchR**:

```{r, eval = FALSE}
library(hatchR)
```

# Built-in parameterizations

**hatchR** ships with built-in model parameterizations, which are included in the `model_table` object. This table includes parameterizations for several salmonid species from hatchery studies relating temperature to hatch and emergence timing [@beacham1990, @austin2019, @sparks2019], which can be selected using `model_select()`.

## `model_table`

`model_table` is loaded with **hatchR**, and is a tibble with 51 rows and 5 columns:

```{r}
model_table
```

-   `author`: the author(s) and year of publication for the model
-   `species`: the species the model is parameterized for
-   `model`: the model ID (if multiple model parameterizations were built (*e.g.*, @beacham1990))
-   `development_type`: the phenological development type (*i.e.*, hatch or emerge)
-   `expression`: a character string defining the model parameterization

## `model_select()`

You will use `author`, `species`, `model`, and `development_type` to select model functions. For instance, if you wanted to access the parameterization for sockeye hatching using model #2 from @beacham1990, you would run:

```{r}
sockeye_hatch_mod <- model_select(
  author = "Beacham and Murray 1990", 
  species = "sockeye", 
  model = 2, 
  development_type = "hatch"
  )

sockeye_hatch_mod
```

Note, that the above arguments are equivalent to the first line and four columns from `model_table`. Your model function object---in this case `sockeye_hatch_mod`---can then be passed to `predict_phenology()`, which we will demonstrate in the [Predict Phenology: Basic](https://bmait101.github.io/hatchR/articles/Predict_phenology_basic.html) vignette.

To see all available characterizations use:

```{r, eval=FALSE}
View(model_table)
```

# Creating custom models

**hatchR** also includes basic functionality to generate your own model parameterizations for predicting hatching and emergence phenology using your own data. **Importantly**, the this functionality implements ***model form 2*** of @beacham1990, which we chose because of its overall simplicity and negligible loss of accuracy. See @beacham1990 and @sparks2019 for more specific discussion regarding model 2 and the development of the effective value approach.

The model follows the general format of:

$$
Effective Value_i = 1/exp(log_ea - log_e(Temperature_i - b))
$$

Where *i* is the daily value and a fish hatches or emerges when the cumulative sum reaches one: $$\sum_{i =1}^nEffectiveValue_i = 1$$

## `fit_model()`

The function `fit_model()` uses data in which average incubation temperature (°C) and days to phenological event are the inputs and parameterizes the above model for *log~e~a* and *b*.

Here, we borrow data from Table 8.1 (pg. 183) from @quinn2018 to generate a custom hatch parameterization for brown trout (*Salmo trutta*).

You could either create a .csv file with those data and import them with `readr::read_csv()` or alternatively, directly input them as an object in R. We'll use `tibble::tibble()` to create the data.

```{r}
bt_data <- tibble(
  temperature = c(2,5,8,11,14), 
  days_to_hatch = c(194,87,54,35,28)
  )
bt_data
```

We can plot our data for a validity check using **ggplot2** (installed with **hatchR**). First load the package:

```{r, eval=FALSE}
library(ggplot2)
```

Now plot the data:

```{r}
bt_data |> 
  ggplot(aes(x = temperature, y = days_to_hatch)) +
  geom_point() +
  theme_classic()
```

We can now use `fit_model()` to create our custom parameterization from our data. You must specify a species and a development type, information which is carried forward in subsequent functions in **hatchR**.

```{r}
bt_fit <- fit_model(temp = bt_data$temperature, 
                    days = bt_data$days_to_hatch, 
                    species = "brown_trout", 
                    development_type = "hatch")
```

The output of `fit_model()` is a list with numerous elements, including the full model, coefficients for *log~e~a* and *b*, the expression you will pass to `predict_phenology()`, a plot to see your fit for your data, and some diagnostic variables. You can see the full extent using:

```{r}
bt_fit
```

The vast majority of the time, what you will want is the actual expression with parameter estimates for use in the `model = ...` argument of `predict_phenology()`. This is stored in the `expression` element of the list, which is a tibble giving the species, development type, and the parameterized model expression.

You can either pass this tibble directly with the `$` operator by calling `$expression` element of the list (e.g., `model = bt_fit$expression`) or set as an object to pass, like so:

```{r, eval=FALSE}
bt_hatch_exp <- bt_fit$expression
bt_hatch_exp
```

`predict_phenology()` will extract the expression from the object and use it to predict phenology. We will demonstrate this in the [Predict phenology: Basic](https://bmait101.github.io/hatchR/articles/Predict_phenology_basic.html) vignette.

## `fit_model()` details

FORTHCOMING.

# Important considerations

Here are some important considerations:

> *Your model fits will only be as good as the data they are generated from*.

1.  We recommend a minimum of four temperature x hatch/emerge data points.

2.  Data should be spread across temperatures as much as possible.

    -   It's much better to have a fit derived from data for temperatures such as 3, 7, 10, 14 °C than it is 8, 9, 10, 11 °C.
    -   The behavior of the model function around the tails of very cold or warm temperatures (relative to the fish species) drive the fit of the function, so more extreme temperatures are helpful.

3.  Think hard about whether the data you are generating your parameterization from match the data from which you are trying to predict or if you are extrapolating beyond what is sensible for the model.

4.  Understand your response variable, most models are fit to 50% hatch or emergence for a family group or population. However, your data may be different and you should interpret your results accordingly (e.g. comparisons between 50% hatch from population A to 95% hatch of population B may not be reasonable).

# References {.unnumbered}
